Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-13 paper.tex
updated text and figures

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo and Fornax clusters to be the
  brightest DM sources and find a particularly low CR induced
  background for Fornax. For a minimum substructure mass given by the
  DM free-streaming scale, the largest DM halo mass maximizes the
  substructure boost over the smoothly distributed annihilation flux:
  for galaxy clusters as the largest collapsed objects, we find a
  boost factor of more than 1000. Since the annihilation flux of
  substructures is mostly contributed by the regions around the virial
  radius, the resulting surface brightness profiles are almost
  flat. This makes it very challenging to detect this flux with
  imaging air Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear
  resolution elements across a cluster; the gamma-ray satellite Fermi
  is better suited. Assuming cold dark matter with a substructure mass
  distribution down to an Earth mass and using Fermi upper limits, we
  rule out the leptophilic models in their present form in 14
  clusters, and limit the boost from Sommerfeld enhancement in Fornax
  to be less than 2 which corresponds to a value in the Milky Way of
  less than 5. Alternatively, if the Sommerfeld enhancement factor is
  realized in Nature, this would imply a limiting substructure mass in
  Fornax of $\mlim > 10^2\,\msun$ which would be a challenge for
  structure formation. In addition, we find that Fermi three-year data
  will be able to constrain the smallest mass of substructures for
  supersymmetric benchmark models or alternatively constrain their
  cross sections for a given substructure boost. The Fermi upper
  limits are closing in on our predictions for the CR flux using an
  analytic model based on cosmological hydrodynamical cluster
  simulations. Thus we will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport
  properties.
\end{abstract}

@@

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like
Segue-1, Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in favour of this possible target for
gamma-ray detection of DM annihilation.  Galaxy clusters constitute
the most massive objects in our Universe that are forming today. This
causes their DM subhalo mass function to be less affected by tidal
stripping compared to galaxy sized halos that formed long ago.  The
annihilation luminosities of the DM halo component for e.g. the Virgo
cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $60$  & $1040$ \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $70$  & $910$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $105$  & $680$  \\
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $130$  & $580$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement. See Section II A 1.\\
  (2) The boost from substructures have been integrated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (5) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric DM
models used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3 we
calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find promising
targets we estimate the fluxes from DM and CRs of all clusters in the
extended HIFLUGCS sample in Section 5. In Section 6 we conclude and
discuss our results.

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um,Aprile:2011ts}). A characteristic of this method is that it
basically only depends on the local DM density (which is rather well
determined to be around $0.4$ GeV/cm$^3$) and the basic cross section
for DM - nucleus scattering. A disadvantage is that it does not benefit
from the two enhancement mechanisms mentioned above, boost from
substructure and/or the Sommerfeld effect. Also, it cannot be excluded
- although it seems at present improbable - that the local halo
structure is such that the solar system happens to be in an underdense
region.

@@

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{1995PhRvD..52.3265A,Hooper:2008kg,Ahlers:2009ae}). In that case,
DM would more likely be explained by the standard scenario of a WIMP,
for example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) caused intense
activity in the theoretical DM community. At first, there was
scepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$\sigv$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/\sigv$.

As LP models, apart from being slightly contrived from the
particle physics point of view, are also rather limited by several
sets of astrophysical data. We take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark LP model that
is still viable.  It is found that the bounds from WMAP5 approximately
imply a Sommerfeld boost factor $S$ which has to satisfy $S(v\to
0)\lesssim 200/f\ (m_Y/1$ GeV$)$, where $f$ is the fraction of energy
from DM annihilation that ionizes the intergalactic medium,
$f\sim 0.7$ for annihilation to electrons and a factor of a few smaller
for all other standard model final states.

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of ($1/2:1/2$)
into ($\mu^+\mu^-:\e^+e^-$).\footnote{This is an approximation to the
  model found in \cite{Finkbeiner:2010sm} that has a branching ratio
  of ($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$).}  The
SFE $S=\B_\sfe(\sigv=200\,\rmn{km}/\rmn{s})\approx 300$ for a MW halo
in our LP model, which we appropriately scale down for cluster halos,
using
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
62.6 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

@@

 Cluster & ID$^{(1)}$ & $F_{\gamma}$$^{(2)}$ & $F_{\gamma}$$^{(2)}$& 
 $F_{\gamma}$$^{(3)}$ & $\eg^2\,\dd F_{\gamma}/\dd \eg$$^{(4)}$ &
 $\eg^2\,\dd F_{\gamma,0.1}/\dd \eg$$^{(4,5)}$ & 
 $\eg^2\,\dd F_{\gamma,1.0}/\dd \eg$$^{(4,6)}$\\

>>>>

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo and Fornax clusters to be the
  brightest DM sources and find a particularly low CR induced
  background for Fornax. For a minimum substructure mass given by the
  DM free-streaming scale, the largest DM halo mass maximizes the
  substructure boost over the smoothly distributed annihilation flux:
  for galaxy clusters as the largest collapsed objects, we find a
  boost factor of more than 1000. Since the annihilation flux of
  substructures is mostly contributed by the regions around the virial
  radius, the resulting surface brightness profiles are almost
  flat. This makes it very challenging to detect this flux with
  imaging air Cherenkov telescopes since their sensitivity drops
  linearly with radius and they typically have $5-10$ linear
  resolution elements across a cluster; the gamma-ray satellite Fermi
  is better suited. Assuming cold dark matter with a substructure mass
  distribution down to an Earth mass and using Fermi upper limits, we
  rule out the leptophilic models in their present form in 26
  clusters, and limit the boost from Sommerfeld enhancement in the
  most constraining cluster M49 to be less than 14. This corresponds
  to a Sommerfeld enhancement in the Milky Way of less than
  8. Alternatively, if the Sommerfeld enhancement factor is realized
  in Nature, this would imply a limiting substructure mass in M49 of
  $\mlim > 10\,\msun$ which would be a challenge for structure
  formation. \C{Furthermore, we find it challenging for Fermi to
    constrain our selection of DM benchmark models using clusters
    without improved modelling of extended sources in addition to a
    stacked likelihood analysis.} The Fermi upper limits are, however,
  closing in on our predictions for the CR flux using an analytic
  model based on cosmological hydrodynamical cluster simulations. Thus
  we will soon start to constrain the underlying CR physics such as
  shock acceleration efficiencies or CR transport properties.
\end{abstract}

@@

2. The dwarf spheroidal galaxies orbiting the Milky Way (MW), like
Segue-1, Ursa Minor, Draco, Sagittarius, Sculptor, Carina or Willman-1
\cite{2009JCAP...01..016B,2010ApJ...720.1174A,2010JCAP...01..031S,2010JCAP...01..031S,2011arXiv1103.0477T,2011APh....34..608H}. The
problem here is that the nature of many of these small, dark
matter-dominated galaxies is not entirely clear, and the velocity
dispersion estimates are based on rather small numbers of stars. Tidal
disruption and confusion with star clusters are other
complications. Thus the DM density profile is very uncertain for most
of them. Nonetheless, by stacking the data together from many dwarf
spheroidals these uncertainties can be made less severe, and
preliminary results from Fermi-LAT shows this method to give quite
promising results \cite{garde}.

3. Galaxy clusters. This possibility has been less studied, however
\C{currently there is an ongoing observational campaign to detect
  gamma-ray emission from galaxy clusters
  \cite{2010ApJ...710..634A,2010JCAP...05..025A,2009A&A...495...27A,2009IJMPD..18.1627D,2009A&A...502..437A,2009arXiv0907.5000G,2009ApJ...704..240K,2009ApJ...706L.275A,2010ApJ...717L..71A}. In fact, we noted} in a previous Letter
\cite{2009PhRvL.103r1302P} that there are certain advantages that work
in favour of this possible target for gamma-ray detection of DM
annihilation. Galaxy clusters constitute the most massive objects in
our Universe that are forming today. This causes their DM subhalo mass
function to be less affected by tidal stripping compared to galaxy
sized halos that formed long ago.  The annihilation luminosities of
the DM halo component for e.g. the Virgo cluster and the Draco dwarf
scales in a way (see \cite{2009PhRvL.103r1302P}) that the ratio of
gamma-ray luminosities from the smooth components is around 4, in
favour of Virgo. In addition, there may be a further enhancement due
to substructure, which to a large extent should be unaffected by tidal
disruption, at least in the outer regions.  According to a recent
estimate \cite{2011MNRAS.410.2309G}, more massive haloes tend to have
a larger mass fraction in subhalos.  For example, cluster size haloes
typically have 7.5 per cent of the mass within $r_{200}$ in
substructures of fractional mass larger than $10^{-5}$, which is 25
per cent higher than for galactic haloes.\footnote{All halo masses and
  length scales in this paper are scaled to the currently favored
  value of Hubble's constant, $H_0 = 70\,
  \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define the virial mass $\mvir$
  and virial radius $\rvir$ as the mass and radius of a sphere
  enclosing a mean density that is 200 times the critical density of
  the Universe $\rho_{\rmn{cr}}$.}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ &  $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ 
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $5.1\times10^{-8}$  & $8.2\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $7.4\times10^{-12}$ & $3.0\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $530/60$  & $1290$ \\
 Perseus \,\,\,\,\,\, & $4.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $6.4\times10^{-12}$ & $2.6\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $530/70$  & $1210$ \\
 Virgo                & $9.2\times10^{-7}$  & $1.4\times10^{-9}$ & $1.1\times10^{-8}$  
 & \,\,\,\,\,         & $1.3\times10^{-10}$ & $5.4\times10^{-12}$ & $3.9\times10^{-12}$ 
 & $6.9$  & $530/105$  & $1110$  \\
 Fornax               & $7.9\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $1.1\times10^{-11}$ & $4.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $530/130$  & $730$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) \C{The mass of Fornax, Coma, and Perseus are taken from \cite{2002ApJ...567..716R}, 
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.}\\
  (2) \C{The boost due to Sommerfeld enhancement. The first value shows the saturated 
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld 
  boost without substructures. See Section II A 1.}\\
  (3) \C{The boost from substructures along the line of sight has been integrated 
    out to an angular size corresponding to $\rvir$.}\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric
benchmark (BM) DM models in this work, and outline the framework
for estimating the gamma-ray emission from various radiative
processes. In Section 3 we calculate the gamma-ray fluxes and spectral
distributions from four clusters and compare it to upper limits set by
Fermi-LAT. For the same clusters we derive the gamma-ray surface
brightness profiles in Section 4. To extend the analysis to more
clusters and find promising targets we estimate the fluxes from DM and
CRs of all clusters in the extended HIFLUGCS sample in Section 5. In
Section 6 we conclude and discuss our results.

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um,Aprile:2011hi}). A characteristic of this
method is that it basically only depends on the local DM density
(which is rather well determined to be around $0.4$ GeV/cm$^3$) and
the basic cross section for DM - nucleus scattering. A disadvantage is
that it does not benefit from the two enhancement mechanisms mentioned
above, boost from substructure and/or the Sommerfeld effect. Also, it
cannot be excluded - although it seems at present improbable - that
the local halo structure is such that the solar system happens to be
in an underdense region.

@@

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{1995PhRvD..52.3265A,Hooper:2008kg,Ahlers:2009ae}). In that case,
DM would more likely be explained by the conventional scenario of a WIMP,
for example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) caused intense
activity in the theoretical DM community. At first, there was
scepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$\sigv$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/\sigv$. \C{Note
  that the $1/v$-scaling is valid only for $(v/c) \gtrsim
  (m_\phi/m_\chi)$; at smaller velocities and outside resonances, the
  $1/v$-enhancement saturates at $m_\phi/m_\chi$
  \cite{2008PhRvL.101z1301K}.}

As LP models, apart from being slightly contrived from the particle
physics point of view, are also rather limited by several sets of
astrophysical data. We take advantage of the recent reanalysis
\cite{Finkbeiner:2010sm} to define a benchmark LP model that is still
viable.  It is found that the bounds from WMAP5 approximately imply a
Sommerfeld boost factor $S$ which has to satisfy $S(v\to
150\,\rmn{km}/\rmn{s})\lesssim 250/f\ (m_\phi/1$ GeV$)$, where $f$ is
the fraction of energy from DM annihilation that ionizes the
intergalactic medium, $f\sim 0.7$ for annihilation to electrons and a
factor of a few smaller for all other standard model final states.

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into
($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). \del{\footnote{This is an
    approximation to the model found in \cite{Finkbeiner:2010sm} that
    has a branching ratio of ($1/4:1/4:1/2$) into
    ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$).}} \C{Furthermore, since most
  of gamma-ray flux is expected from come from dense subhalos at which
  the velocity dispersion is close to the velocity limit $v_\rmn{sat}$
  where the SFE saturates, we use $S=\B_\sfe(\sigv=v_\rmn{sat})\approx
  530$ \footnote{Note that, $S$ is strictly speaking the Sommerfeld
    enhanced annihilation cross-section, but here we have in addition
    included the order unity constraint that make sure that the relic
    density is not over-depleted at freeze-out.} for the cluster
  halos. However, for those cluster halos without a boost from
  substructures, we adopt a velocity dependent SFE that is normalized
  to fit the electron and positron excess at Earth observed by the
  PAMELA and Fermi satellites,}
\begin{equation}
\B_\sfe(\sigv) = 0.266\, \frac{c}{v_\rmn{sat}}\, \frac{v_\rmn{MW}}{\sigv} = 
121 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\label{eq:B_sfe}
\end{equation}
Here the velocity dispersion for the MW $v_\rmn{MW} \approx
220\,$~km/s and the velocity dispersion $\sigv$ for clusters is
derived in \cite{2005RvMP...77..207V}. Clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@
